\documentclass[12pt,fleqn]{article}

%\setlength{\parindent}{2em}
\setlength{\parskip}{0ex}
\setlength{\oddsidemargin}{0in}
\setlength{\textwidth}{6.5in}
\setlength{\topmargin}{-.5in}
\setlength{\textheight}{9in}

\newcommand{\comment}[1]{}
\renewcommand{\thepage}{}

\begin{document}

\setcounter{page}{1}
\thispagestyle{empty}

\begin{center}
\LARGE
Homework 2 \\
\normalsize 
\ \\
CS 2233/2231 -- Fall 2012 \hfill assigned September 6, 2012 \\
Tom Bylander, Instructor \hfill due September 17, 2012
\end{center}

\normalsize

\begin{enumerate}

\item (15 points) Section 1.4: 8 (no more than 10 words each), 12,
  16.

\item (5 points) Section 1.5: 2

\item (20 pts.)
Determine the truth value of each statement for each domain.

\begin{center}
\begin{tabular}{l|l|l|l|l}
 & \multicolumn{4}{c}{Domain} \\
 & Real Numbers & Positive Reals & Integers & Positive Integers \\ \hline
$\exists x (x = -x)$ & & & & \\
$\forall x (2x \leq 3x)$ & & & & \\
$\exists x (x^2 = 2)$ & & & & \\
$\forall x (x \leq x^2)$ & & & & \\
$\forall x \exists y (xy = 1)$ & & & &
\end{tabular}
\end{center}

\item (10 pts.) Express the negations of each of these statements so
  that there are no negation symbols.  For the last statement, recall
  that $p \rightarrow q$ is equivalent to $\neg p \vee q$.

\begin{enumerate}
\item $\forall x (x^2 > 0)$
\item $\exists x (x^2 < 0)$
\item $\forall x (x < 0 \vee x > 0)$
\item $\exists x (x < 0 \wedge x > 0)$
\item $\forall x (x < 0 \rightarrow x > 0)$
\end{enumerate}

\item (10 points) Determine the truth value of each of these
  statements.  The domain is the real numbers.

\begin{enumerate}
\item $\exists x (x < x^2)$

\item $\forall y (y < y^2)$

\item $\exists x \forall y (x < y^2)$

\item $\forall y \exists x (x < y^2)$

\item $\exists y \forall x (x < y^2)$

\item $\forall x \exists y (x < y^2)$

\item $\exists x \forall y (x > y^2)$

\item $\forall y \exists x (x > y^2)$

\item $\exists y \forall x (x > y^2)$

\item $\forall x \exists y (x > y^2)$
\end{enumerate}

\item (5 points) Show that $\neg \exists x \, \neg P(x)$ is equivalent
to $\forall x \, P(x)$.

\item (5 points) Prove that disjunctive syllogism is a rule of inference.

\item (5 points) Use a truth table to prove that $q$ can be inferred
  from $p \rightarrow q$ and $\neg p \rightarrow q$.  This could be thought
  of as the rule of inference underlying proof by cases.

\item (5 points) What combinations of the following are sufficient to
conclude $x < 0$?

\begin{itemize}
\item $y < 0$
\item $y = 0$
\item $y > 0$
\item $x < y$
\item $x > y$
\item $x = y$
\end{itemize}

\item (10 points) Suppose $x > 1$. Which of the following are valid
  conclusions?  Provide a brief justification of each answer.

\begin{enumerate}
\item $x > 0$
\item $x > 2$
\item $x$ is even.
\item $x > -x$
\item $y > 1 \vee y < x$
\end{enumerate}


\item (10 points) Suppose $x + y$ is an even number and $x + y > 0$.
  Which of the following are valid conclusions?  Provide a brief
  justification of each answer.

\begin{enumerate}
\item $x > 0 \wedge y > 0$
\item $x > 0 \vee y > 0$
\item $x > 0 \oplus y > 0$
\item $x \mbox{\ is odd\ } \leftrightarrow y \mbox{\ is odd}$
\item $x \mbox{\ is even\ } \vee y \mbox{\ is odd}$
\end{enumerate}

\end{enumerate}

\end{document}
